---
title: "Impact of sever weather events on public health and economic of the U. S. between 1950 and 2011"
author: "Valentine Savchenko"
date: "Saturday, May 23, 2015"
output: html_document
---
### Synopsis
The report gives brief summary on impact of sever weather events on public and economic of the U. S. between 1950 and 2011.

### Initial analysis of data
```{r, setoptions, echo = F, warning = F}
library(knitr)
opts_chunk$set(echo = T)
opts_chunk$set(warning = F)
opts_chunk$set(message = F)
```
The data was obtained from the U.S. National Oceanic and Atmospheric Administration's (NOAA) [storm database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
```{r}
stormZip <- "storm.bz2"
if (file.access(stormZip, 4))
{
  stormUrl <-
    "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(stormUrl, stormZip)
}
storm <- read.csv(stormZip, stringsAsFactors = F)
```
The total number of observations in the set is
```{r}
nrow(storm)
```
#### Event types
According to [NWS Directive 10-1605](http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf) any event must fall into one of following categories
```{r}
eventTypes <-
  c("ASTRONOMICAL LOW TIDE", "AVALANCHE",               
    "BLIZZARD",
    "COASTAL FLOOD", "COLD/WIND CHILL", "DEBRIS FLOW",
    "DENSE FOG", "DENSE SMOKE", "DROUGHT", "DUST DEVIL", "DUST STORM",
    "EXCESSIVE HEAT", "EXTREME COLD/WIND CHILL",
    "FLASH FLOOD", "FLOOD", "FROST/FREEZE", "FUNNEL CLOUD", "FREEZING FOG", 
    "HAIL", "HEAT", "HEAVY RAIN", "HEAVY SNOW", "HIGH SURF", "HIGH WIND", "HURRICANE/TYPHOON",
    "ICE STORM",
    "LAKE-EFFECT SNOW", "LAKESHORE FLOOD", "LIGHTNING",
    "MARINE HAIL", "MARINE HIGH WIND", "MARINE STRONG WIND", "MARINE THUNDERSTORM WIND",
    "RIP CURRENT",
    "SEICHE", "SLEET", "STORM SURGE/TIDE", "STRONG WIND",
    "THUNDERSTORM WIND", "TORNADO", "TROPICAL DEPRESSION", "TROPICAL STORM", "TSUNAMI",
    "VOLCANIC ASH",
    "WATERSPOUT", "WILDFIRE", "WINTER STORM", "WINTER WEATHER")
```
The propotion of properly marked observations in the set is
```{r}
properlyMarked <- toupper(storm$EVTYPE) %in% eventTypes
sum(properlyMarked) / length(storm$EVTYPE) * 100
```
Following events either aren't represented in the set or improperly marked
```{r}
eventTypes[!(eventTypes %in% storm$EVTYPE)]
```
The propotion of partially marked observations in the set is
```{r}
partiallyMarked <- storm$EVTYPE %in%
                   grep(paste(eventTypes, collapse = "|"),
                        unique(storm$EVTYPE[!properlyMarked]), value = T)
sum(partiallyMarked) / length(storm$EVTYPE) * 100
```
#### Economic damage
The set contains two kinds of economical losts from severe weather events. The first one is property damage, which is divided between two columns `PROPDMG` and `PROPDMGEXP`. According to paragraph 2.7 of [NWS Directive 10-1605](http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf). `PROMPDMG` must contain three significant digits. The proportion of observations that obey this recommendation is
```{r}
length(grep("^\\d\\.\\d\\d$", format(storm$PROPDMG, digits = 2, trim = T))) / length(storm$PROPDMG) * 100
```
`PROPDMGEXP` must contain alphabetical characters used to signify magnitude: "K" for thousands, "M" for millions, and "B" for billions. The proportion of observations that follow it is
```{r}
length(grep("K|M|B", storm$PROPDMGEXP)) / length(storm$PROPDMGEXP) * 100
```
Magnitude characters actually used in the set
```{r}
library(dplyr)
storm %>% group_by(PROPDMGEXP) %>% summarize(number = n())
```
Situation with the crop damage is quite similar
```{r}
length(grep("^\\d\\.\\d\\d$", format(storm$CROPDMG, digits = 2, trim = T))) / length(storm$CROPDMG) * 100
storm %>% group_by(CROPDMGEXP) %>% summarize(number = n())
```
"H" and "h" alphabetic characters could be interpret as magnitude of 10^2. "m" - as simply "M". "0"-"8" - as 10^0, 10^1 etc. "", "+", "-" and "?" as 10^0.
### Data processing
The impact on public health is judged based on numbers of killed `FATALITIES` and injuried `INJURIES` people. The economic impact - on property damage `PROPDMG` and crop damage `CROPDMG` in billions of U. S. $.
```{r}
storm <- storm %>% select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
storm$EVTYPE <- toupper(storm$EVTYPE)
for (i in 1:length(eventTypes))
{
  storm$EVTYPE[grep(paste0("^", eventTypes[i], "$"), storm$EVTYPE)] <- eventTypes[i]
}
storm$EVTYPE[!(storm$EVTYPE %in% eventTypes)] <- "ANY"

storm$PROPDMGEXP[storm$PROPDMGEXP %in% c("B")] <- "9"
storm$PROPDMGEXP[storm$PROPDMGEXP %in% c("m", "M")] <- "6"
storm$PROPDMGEXP[storm$PROPDMGEXP %in% c("K")] <- "3"
storm$PROPDMGEXP[storm$PROPDMGEXP %in% c("h", "H")] <- "2"
storm$PROPDMGEXP[storm$PROPDMGEXP %in% c("", "+", "-", "?")] <- "1"
stopifnot(sum(is.na(as.numeric(storm$PROPDMGEXP))) == 0)

storm$CROPDMGEXP[storm$CROPDMGEXP %in% c("B")] <- "9"
storm$CROPDMGEXP[storm$CROPDMGEXP %in% c("m", "M")] <- "6"
storm$CROPDMGEXP[storm$CROPDMGEXP %in% c("k", "K")] <- "3"
storm$CROPDMGEXP[storm$CROPDMGEXP %in% c("", "?")] <- "1"
stopifnot(sum(is.na(as.numeric(storm$CROPDMGEXP))) == 0)

for (i in 0:9)
{
  ii <- as.character(i)
  storm$PROPDMG[storm$PROPDMGEXP == ii] <- storm$PROPDMG[storm$PROPDMGEXP == ii] * (10^i)
  storm$CROPDMG[storm$CROPDMGEXP == ii] <- storm$CROPDMG[storm$CROPDMGEXP == ii] * (10^i)
}

storm <- storm %>% select(EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG)
```
### Results
Event types ordered by total number of people killed
```{r}
healthImpact <- storm %>%
                group_by(EVTYPE) %>%
                rename(Type = EVTYPE) %>%
                summarise(Killed = sum(FATALITIES),
                          Injuried = sum(INJURIES)) %>%
                arrange(desc(Killed), desc(Injuried))
healthImpact
```
Even though tornado's impact on public health have been recorded for more than 60 years, just 15 years of collecting data on the heat-related events shows that this is quite serious problem. It probably was overlooked in the past due to inability to effectively collect data.

Event types ordered by the total property damage in bilions of U. S. dollars
```{r}
economicImpact <- storm %>%
                group_by(EVTYPE) %>%
                rename(Type = EVTYPE) %>%
                summarise(Property = sum(PROPDMG),
                          Crop = sum(CROPDMG)) %>%
                arrange(desc(Property), desc(Crop))
economicImpact
```
Just 15 years of observations shows that floods entail the largest economic consequences. They also kill significant number of people.

The figure below summarize total economic impact of the top 10 most sever weather event types
```{r}
library(ggplot2)

totalDamage <- sum(storm$PROPDMG) + sum(storm$CROPDMG)
relativeEconomicImpact <- storm %>%
                          group_by(EVTYPE) %>%
                          rename(Type = EVTYPE) %>%
                          summarise(Damage = sum(PROPDMG) + sum(CROPDMG)) %>%
                          top_n(10, Damage) %>%
                          arrange(desc(Damage))
relativeEconomicImpact$Damage <- relativeEconomicImpact$Damage / totalDamage * 100
g <- ggplot(relativeEconomicImpact, aes(x = "", y = Damage, fill = Type)) +
     geom_bar(width = 1, stat = "identity") +
     ggtitle("Total relative economic impact of the top 10 most severe weather event types")
print(g)
```

### Further work
One may notice that improperly marked events ("ANY") killed quite a few people and caused significant amount of damage, so more effort should be made to properly classify observations in NOAA's set. This work is planned for the future versions of the report.
